---
title: "My R Package: knntuner"
---

## Overview

`knntuner` is an R package I developed for my Advanced Data Computing course. The package is designed to streamline parts of the K-nearest neighbors modeling workflow in `caret` by helping users create tuning grids, define cross-validation settings, and identify the best-performing value of \(k\).

The goal of the package is to make KNN tuning more efficient and accessible for students learning statistical machine learning in R, especially for students in STATS 205 - Statistical Machine Learning at Lawrence University. Please see the Github Repo and package website (below!) for more information.

## Package Purpose

K-nearest neighbors models often require repeated code for cross-validation and tuning different values of \(k\) when working with the `caret` package. `knntuner` helps simplify this workflow through a small set of helper functions.

## Main Functions

- `kgrid()` creates a data frame of candidate \(k\) values for model tuning.
- `cv_control()` creates cross-validation specifications using `caret`.
- `bestk()` identifies the best-performing \(k\) value from a trained model.

## Example Workflow

The package includes a simulated student performance dataset called `student_knn_data`. Below, I go through a sample workflow using `knntuner` and its dataset `student_knn_data`.

In the example, make sure to run each of the code chunks in the same order that they are presented below - they represent the basic statistical machine learning workflow. Please see the vignette on the package website for more detailed information on each package function, and ways to customize their outputs.

### Installation

You can install the development version of `knntuner` from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
# install.packages("devtools")
devtools::install_github("ADC-405-S26/knntuner")
```

### Workflow

This is a basic example which shows how `knntuner` can be used in the STATS 205 KNN model building and tuning workflow.

```{r example, message = FALSE}
library(knntuner)
library(rsample)
library(caret)
library(recipes)
# necessary packages
```

In this package, there is a small simulated dataset of 100 observations in 6 variables. The data measure student characteristics, and track their performance. A head of the dataset can be seen below, and further information on each variable can be found in the vignette on the package website.

```{r}
head(student_knn_data) # simulated dataset as a part of knntuner (100 observations)
```


First, start with the standard split of data into a training set and a test set. This example uses the student dataset built into the package.

```{r}
set.seed(208)

knntuner_data_split <- initial_split(data = student_knn_data, prop = 0.8) # setting training data to be 80%, and test data to be 20%, and saving these settings as an object

student_train <- training(knntuner_data_split) # training data
student_test  <- testing(knntuner_data_split) # test data
```

Next, create a recipe to remove the ID column, dummy-code categorical predictors, and scale the numeric predictors for KNN, to prepare the data for analysis.

```{r}
student_recipe <- recipe(pass ~ ., data = student_train) |>
  step_rm(id) |> # ID will not be used as a predictor
  step_dummy(all_nominal_predictors()) |> # dummy encoding nominal categorical variables
  step_center(all_numeric_predictors()) |> # normalizing numeric values step 1
  step_scale(all_numeric_predictors()) # normalizing numeric values step 2
```


Then, define your cross-validation specifications with the `cv_control` function.

```{r}
cv <- cv_control() # creating the cv specifications and storing them as cv
```

Next, you can train and cross-validate your model using the specifications that you set. Firstly, use the `kgrid` function to quickly create a dataframe to hold the cross-validation results for each k value. `kgrid` allows you to choose the minimum and maximum k values to test, as well as the step size to take.

```{r}
set.seed(208)

k_grid <- kgrid(1, 9, 2) # knntuner function here, creating the desired kgrid and storing it as an object

student_model <- train(
  student_recipe,
  data = student_train,
  method = "knn",
  trControl = cv, # knntuner object here!
  tuneGrid = k_grid, # knntuner object here!
  metric = "Sens"
)
```

Finally, use the `bestk` function to identify the best performing k value to choose for your final model! Now, you can move on to training your final model, and analyzing its performance on the test data.

```{r}
bestk(student_model) # finding the optimal k value!
```



## Project Materials

- [View GitHub Repository](https://github.com/ADC-405-S26/knntuner)
- [View Package Website](https://adc-405-s26.github.io/knntuner/)